Certifying, incremental type checking

### 2012-06-16 doc/papers/LFMTP2012/paper.tex
paper: dernieres corrections

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for mere ``type checking'', or explicit
``church-style'' calculi. Take for example dependent type systems,
relying definitional equality on types, or type systems with subtype
polymorphism: both rely on a rule of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer has to turn the declarative system into an algorithm, and
convince himself that both are equivalent. This task is highly
non-trivial and non-modular: for instance, how to infer the type of a
System \sysname{T} recursor $\recb M N x y P$, in presence of
subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it does not have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
inference rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or return an
incorrect derivation. Explicitly generating the derivation in the
declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a well-typed program in one try, compile it
and run it: writing a program longer than a few dozens of lines
becomes a tight and non-linear interaction between the human and the
type-checker involving constant experiments, fixes, etc. If this
interaction is possible for short programs and fast checkers, it
becomes increasingly hindered by the latency of complete rechecks:
even if the modifications made to the program between two interactions
are small, the whole program is usually rechecked. This is especially
true for languages embedding formal verification aspects, like proof
assistants: there, tight interaction with a type or proof checker is
unavoidable due to the difficulty of writing correct proofs without
guidance from the system, and the time taken by many tools to infer
easy parts of the proofs (proof search). The constant modification of
the source makes it necessary to observe \emph{incrementally} the
effect of a small change on the whole edifice: we need to take
advantage of the results of previous checks in order to build the
result for a modified version of the program.

Several generic methods have been devised for turning a batch program
into an incremental one \cite{acar2006,carlsson2002}. Unfortunately,
they suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

Now, if we consider as above a type-checker as a derivation-generating
procedure, and storing these derivations between calls to the
procedure, we could recognize already-built subderivations for
subterms common to an old, checked version and a new, to-be-checked
version, and graft together these pieces of derivation to form a new,
large one for the new version of the program, saving the regeneration
of all equal subterms' derivations in compatible contexts. This way we
constantly maintain the verifiable witness of well-typing for the
program, and take advantage of the quotients made on higher-order
syntax.

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on these types (for
formal details, see appendix). He (or she) wants to write a program
computing $(2+2)\times 3$ in that system. He knows that sometimes the
arguments of $\cst{rec}$ are quite hard to get correct, but
\sysname{T_{<:}} is a typed language; he wants to take advantage of
this to build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1 =
\recb{\s\z}{\s\z} x y {\s {\var x}}$: he sends the command
$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
well-formed program and tries to construct a valid typing derivation
for its argument in the current, empty context. The system succeeds
and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
\cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$. He
could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program: this is a
\emph{delta}. Derivation $\md_2$ used to live in the empty context, we
are then free to use it in the new context where
$\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T
enjoys the \emph{weakening} property; $\md_4$ depending on a
derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole with
the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them.

@@

We choose here to represent deltas as usual terms with variables $\var
x, \var y$ and special, \emph{meta}-variables referring to
already-computed results $\md_i$. Commands sent to the system consists
of terms, possibly containing special typed operators having a
computational content: $\finfer{\cdot}$ takes a term $M$ and produces
the (dependent) pair of a type $A$ and a derivation of the judgment
$\vdash M:A$ in the current context; $\fget{\cdot}$ does the exact
converse: it takes a derivation $\vdash M:A$ and projects it back as
the term $M$. The successful result of a command execution is a
derivation where each subderivation has been given a unique
metavariable name for later reuse (a \emph{repository}). As
subderivations can have free variables, we close them by a local
environment that must be instantiated by a \emph{substitution}.

Although this small example may seem anecdotal, the interest of such a
mechanism becomes clear on very large programs: then the cost of
rechecking well-typing of a small modification amounts to rechecking
the path in the term from the root to the changed part, and the
changed part itself.

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques \cite{acar2006}. However,
it is best understood in the case of purely first-order data, and to
our best knowledge remains unexplored for higher-order structures
containing binders, like programs and proofs.

@@

\noindent
We write $\pair{\cdot}{\cdot}$ for the constructor of $\Sigma$-types.
Patterns do not have to be exhaustive since we are in an untrusted
setting. Fr conciseness, we use case failure for signaling typing
error. There are three cases for $\s{M}$ in the system, corresponding
to the three rules in the declarative system. For the application
case, we rely on the well-known property that we may need to insert a
coercion from the actual type of the argument to its expected type:

@@

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error. Notice the slight optimization to the size of the derivation:
if the types $A_1$ and $A_2$ are the same syntactically, we do not
generate a subtyping rule. This is done by matching on the returned
derivation $\md_\leq$ in the case of the \textsc{Refl} rule.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$. Otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
($\match M \ldots$ being syntactic sugar for $\matchin M \envnil
\ldots$). Even then, in the recursive call, we would have no way of
synthesizing a variable $\var x$'s type anymore. Thus we substitute
all occurrences of $\var x$ with the \emph{output} of function
$\finfer{}$ on it (since it is known at this point), \ie\ a pair of
$A$ and $\md_1$. For this substitution to be well-typed, we now need
to coerce this output to a valid input (a term). We introduce an
operator which is in some sense the \emph{inverse} of $\finfer{}$:
operator $\fget{}$ maps a pair of a type and a derivation to the term
it types.

@@

One problem remains with subtyping: terms $N : A_N$ and $P : A_P$
should be coercible to the same type $A$, but $P$ should be typed
under $\vdash y:A$, which is not known before typing $P$. A trick to
break the loop is to iterate the typing of $P$ with a more and more
general type, until we reach a fixpoint. We reach a fixpoint if there
is a unique more general type, in our particular case of subtyping
relation, this means iterating only twice: we will then attain
\cst{nat}.

@@

Note that $\fget{}$ can be derived automatically from the type of
$\finfer{}$: we will call it its inverse, since it maps the return
type to the arguments. It has definition:

@@

The idea of untrustedly generating certificates verified \emph{a
  posteriori} is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: they are untrusted programs generating pieces of
proofs, assembled by other, higher-order tactics; verification of the
generated proofs is done when the evaluation of all tactics is
complete. The differences with our mechanism are two-fold: first,
tactics usually take one (or several) \emph{goal}(s) as input, \ie\
holes in a proof, whereas our functions can take any term; secondly,
proof-check is usually done at the end of the evaluation, which can
lead to less tractable bugs; by making evaluation and checking
mutually recursive, we choose to detect errors earlier in these
witnesses: exactly at the point where they are generated.

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special term construct
$\{x:A\}$ denoting free variables is added to the syntax of terms, and
rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a dependently typed lambda-calculus for \emph{representing}
higher-order terms and derivations
\cite{pfenning1988higher}. The user gives a
\emph{signature} defining abstract syntax and inference rules of his
\emph{object language}, and gets an algorithm to verify terms and
derivations expressed in the \LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}

To write deltas by sharing subterms, we need to be able to refer to
any subterm. For this, we first extend \LF\ with contextual
metavariables (written $\meta X$, $\meta Y$) standing for open terms,
following \cite{nanevski2008contextual}, secondly, we make sure that
every subterm is named by a metavariable (an operation called
\emph{slicing}). We thus introduce Sliced \LF\ (\SLF, see
Fig. \ref{fig:syntax-LF}): it is an extension of \emph{spine canonical
  \LF} \cite{pfenning2007term} enriched with metavariables which are
references to open terms as in \cite{nanevski2008contextual}.

@@

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence. Canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition of substitution and
$\eta$-long forms, and corresponds to a focused sequent calculus known
as \sysname{LJT}~\cite{herbelin1995lambda}. We differ the discussion
of function symbols $\fct f$ to section \ref{sec:comput}. A
\emph{value} is an object containing no function symbols. The
definition of hereditary substitution $\gsubst M \sigma$ is skipped
and can be found in the appendix.

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}. Canonical objects
are checked against a given type, whereas we can synthesize the type
of atomic objects; the coercion from canonical to atomic objects
triggers the equality check between synthesized and checked types.

@@

If $\mr = (\Delta, X[\sigma])$, we write $\hat {\mr}$ for the tip
$X[\sigma]$, abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$
for $\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
The algorithm in \ref{sec:eval} will maintain two invariants: a
repository $\mr$ is \emph{acyclic}, (for all $\meta X$ in
$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta
X$) and all $F$ and $P$ in it are values.

>>>>

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for mere ``type checking'', or explicit
``church-style'' calculi. Take for example dependent type systems,
relying on definitional equality on types, or type systems with subtype
polymorphism: both feature a rule of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer has to turn the declarative system into an algorithm, and
convince himself that both are equivalent. This task is highly
non-trivial and non-modular: for instance, how to infer the type of a
System \sysname{T} recursor $\recb M N x y P$, in presence of
subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path: derivations in the
declarative systems, if they cannot always be easily inferred, can be
\emph{verified} if provided by an oracle. We can tell cheaply if it is
wrong by verifying the witness typing derivation it provided. Let us
then imagine a type-checker to be the pair of \emph{(i)} an untrusted
procedure $\finfer M$ to infer a typing derivation out of program $M$;
it does not have to be complete, or even sound, and \emph{(ii)} a
trusted and fast kernel $\fcheck{\md}$ to verify these derivations
\wrt\ a given set of inference rules $\Sigma$. This is a
\emph{certifying} scheme (see \cite{leroy2006compcert}). If $\fcheck{\finfer
  M} = \cst{true}$ then there really \emph{is} a derivation $\md$ for
$M$, but the converse is not necessarily true: function
$\finfer{}\!\!$ could fail, or return an incorrect
derivation. Explicitly generating the derivation in the declarative
system has the following advantages over the first approach:
\begin{itemize}
\item It is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof.
\item Compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item It inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

As type systems become richer, it becomes increasingly difficult for
the programmer to write a well-typed program in one try, compile it
and run it: writing a program becomes a tight and non-linear
interaction between the human and the type-checker involving constant
experiments, fixes, etc. If this interaction is possible for short
programs and fast checkers, it becomes increasingly hindered by the
latency of complete rechecks: even if the modifications made to the
program between two interactions are small, the whole program is
usually rechecked. This is especially true for languages embedding
formal verification aspects, like proof assistants: there, tight
interaction with a type or proof checker is unavoidable due to the
difficulty of writing correct proofs without guidance from the system,
and the time taken by many tools to infer easy parts of the proofs
(proof search). The constant modification of the source makes it
necessary to observe \emph{incrementally} the effect of a small change
on the whole edifice: we need to take advantage of the results of
previous checks in order to build the result for a modified version of
the program.

Several generic methods have been devised for turning a batch program
into an incremental one
\cite{pugh1989incremental,acar2006,carlsson2002}. Unfortunately, they
suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item First, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still
  valid when put under a binder $\lam y (x + 2) - y$ and should not be
  regenerated.
\item Since it is easy to track edits on a very large programs (trace
  of the editor), we may want to feed the type checker only with a
  small \emph{delta} on the previous version and not the whole
  program, thus saving the time to recognize already checked parts;
\item Finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

Now, if we consider as above a type-checker as a derivation-generating
procedure, and storing these derivations between calls to the
procedure, we can reuse already-built subderivations for
subterms common to an old, checked version and a new, to-be-checked
version, and graft together these pieces of derivation to form a new,
large one for the new version of the program, saving the regeneration
of all equal subterms' derivations in compatible contexts. This way we
constantly maintain the verifiable witness of well-typing for the
program, and take advantage of the quotients made on syntax with
binders.

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on these types. He (or
she) wants to write a program computing $(2+2)\times 3$ in that
system. He knows that sometimes the arguments of $\cst{rec}$ are quite
hard to get correct, but \sysname{T_{<:}} is a typed language; he
wants to take advantage of this to build his program incrementally
using the teacher's provided incremental verifier. As a first attempt,
he wants to verify $P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: he
sends the command $\finfer{P_1}$ to the system. The $\finfer{}$
operator takes a well-formed program and tries to construct a valid
typing derivation for its argument in the current, empty context. The
system succeeds and returns a typing derivation $\md_1$ of \/ $\vdash
P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$. He
could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program: this is a
\emph{delta}. Derivation $\md_2$ used to live in the empty context, we
are then free to use it in the new context where
$\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T
enjoys the \emph{weakening} property; $\md_4$ depending on a
derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole with
the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates the derivation for this application, it inserts a
coercion $\cst{odd} \leq \cst{nat}$ at these two points. Taking
advantage of $\md_2$ and $\md_4$ the system succeeds with a new
derivation made out of these, without having to regenerate them.

@@

We choose here to represent deltas as usual terms with variables $\var
x, \var y$ and special, \emph{meta}-variables referring to
already-computed results $\md_i$. Commands sent to the system consists
of terms, possibly containing special typed operators having a
computational content: $\finfer{}$ takes a term $M$ and produces the
(dependent) pair of a type $A$ and a derivation of the judgment
$\vdash M:A$ in the current context; $\fget{}$ does the exact
converse: it takes a derivation $\vdash M:A$ and projects it back as
the term $M$. The successful result of a command execution is a
derivation where each subderivation has been given a unique
metavariable name for later reuse (a \emph{repository}). As
subderivations can have free variables, we close them by a local
environment that must be instantiated by a \emph{substitution}.

Although this small example may seem anecdotal, the interest of such a
mechanism becomes clear on very large programs: then the cost of
rechecking well-typing of a small modification amounts to recheck the
path in the term from the root to the changed part, and the changed
part itself.

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subresults of the output $\md_i$ back into the
input $P_i$ is a well-known technique for handling changes in the
input: combined with a way to automatically recognize already-computed
input, it is \emph{memoization}, a general-purpose technique for
incremental computation that has been extensively studied and combined
with other techniques \cite{acar2006,pugh1989incremental}. However, it
is best understood in the case of purely first-order data, and to our
best knowledge remains mainly unexplored for higher-order structures
containing binders, like programs and proofs.

@@

\noindent
We write $\pair{\cdot}{\cdot}$ for the constructor of $\Sigma$-types.
Patterns do not have to be exhaustive since we are in an untrusted
setting. For conciseness, we use case failure for signaling typing
error. There are three cases for $\s{M}$ in the system, corresponding
to the three rules in the declarative system. For the application
case, we rely on the well-known property that we may need to insert a
coercion from the actual type of the argument to its expected type:

@@

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error. Notice the slight optimization to the size of the derivation:
if the types $A_1$ and $A_2$ are the same syntactically, we do not
generate a subtyping rule. This is done by matching on the returned
derivation $\md_\leq$, in the case of the \textsc{Refl} rule.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$. Otherwise the check that the open
derivation $\md$ returned by the recursive call is correct would fail
($x$ is unknown). We thus introduce a new construct $\matchin M \Gamma
\ldots$ ($\match M \ldots$ being syntactic sugar for $\matchin M
\envnil \ldots$). Even then, we would have no way of synthesizing a
variable $\var x$'s type anymore ($\caseb{\var x} \text{???}$). We
therefore substitute all occurrences of $\var x$ with the
\emph{output} of function $\finfer{}$ on it (since it is known at this
point), \ie\ a pair of $A$ and $\md_1$. For this substitution to be
well-typed, we now need to coerce this output to a valid input (a
term). We introduce an operator which is in some sense the
\emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
type and a derivation to the term it types.

@@

One problem remains with subtyping: terms $N : A_N$ and $P : A_P$
should be coercible to the same type $A$, but $P$ should be typed
under $\vdash y:A$, which is not known before typing $P$. A trick to
break the loop is to iterate the typing of $P$ with a more and more
general type, until we reach a fixpoint. We reach a fixpoint if there
is a unique more general type, in our particular case of subtyping
relation, this means iterating only twice: we will then reach
\cst{nat}.

@@

Note that $\fget{}$ can be derived automatically from the type of
$\finfer{}$: we will call it its inverse, since it maps the return
type to the argument. It has definition:

@@

The idea of untrustedly generating certificates verified \emph{a
  posteriori} is well-known to \sysname{LCF}-style proof assistants:
during manual proof search, the interaction is driven by
\emph{tactics}: they are untrusted programs generating pieces of
proofs, assembled by other, higher-order tactics; verification of the
generated proofs is done when the evaluation of all tactics is
complete. The differences with our mechanism are two-fold: first,
tactics usually take one (or several) \emph{goal}(s) as input, \ie\
holes in a proof, whereas our functions can take any term; secondly,
proof-check is usually done at the end of the evaluation, which can
lead to less tractable bugs; by making evaluation and checking
mutually recursive, we choose to detect errors earlier in these
witnesses: exactly at the point where they are generated.

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special term construct
$\{x:A\}$ denoting free variables is added to the syntax of terms, and
rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:eval}).

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solutions to this question. It
is a dependently typed lambda-calculus for \emph{representing}
higher-order terms and derivations \cite{pfenning1988higher}. The user
gives a \emph{signature} defining abstract syntax and inference rules
of his object language, and gets an algorithm to verify terms and
derivations expressed in the \LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}

To write deltas by sharing subterms, we need to be able to refer to
any subterm. For this, we first extend \LF\ with contextual
metavariables (written $\meta X$, $\meta Y$) standing for open terms,
following \cite{nanevski2008contextual}. Secondly, we make sure that
every subterm is named by a metavariable (an operation called
\emph{slicing}). We thus introduce Sliced \LF\ (\SLF, see
Fig. \ref{fig:syntax-LF}): it is an extension of \emph{spine canonical
  \LF} \cite{pfenning2007term} enriched with metavariables which are
references to open terms as in \cite{nanevski2008contextual}.

@@

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence. Canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition of substitution and
$\eta$-long forms, and corresponds to a focused sequent calculus known
as \sysname{LJT}~\cite{herbelin1995lambda}. We differ the discussion
of function symbols $\fct f$ to section \ref{sec:comput}. A
\emph{value} is an object containing no function symbols. The
definition of hereditary substitution $\gsubst M \sigma$ is standard,
and it is eluded here.

We distinguish syntactically between \emph{checkable} and
\emph{inferrable} objects (resp. \emph{canonical} $M$ and
\emph{atomic} $F$), as in \eg\ \cite{pierce2000local}. Canonical objects
are checked against a given type, whereas we can synthesize the type
of atomic objects; the coercion from canonical to atomic objects
triggers the equality check between synthesized and checked types.

@@

If $\mr = (\Delta, X[\sigma])$, we write $\hat {\mr}$ for the tip
$X[\sigma]$, and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$ if
$\Delta(\meta X) = (\Gamma\vdash M : A)$.  The algorithm in
\ref{sec:eval} will maintain two invariants: a repository $\mr$ is
\emph{acyclic}, (for all $\meta X$ in $\mr$, the fixpoint of
$\FMV{\mr(\meta X)}$ does not contain $\meta X$) and all $F$ and $P$
in it are values.
